---
title: "EasyDE Walkthrough — 00 Setup"
output: html_document
---

# Setup

This notebook loads libraries, reads the config, and defines the contrast ×
stratum loops used by all subsequent notebooks.

**Run this notebook first.**

## Set working directory

Change this to **your pipeline root** (the directory containing `Snakefile`).

```{r set-root}
# EDIT THIS: path to your EasyDE pipeline root
pipeline_root <- normalizePath("../..")
knitr::opts_knit$set(root.dir = pipeline_root)
setwd(pipeline_root)
cat("Working directory:", getwd(), "\n")
```

## Choose your config

```{r config-choice}
# EDIT THIS: pick your config file
config_path <- "config/config_treatments.yaml"
```

## Load libraries and utilities

```{r load-libs, message=FALSE, warning=FALSE}
suppressMessages({
    library(yaml)
    library(data.table)
    library(dplyr)
    library(ggplot2)
})

# Source all pipeline utilities
for (f in list.files("workflow/scripts/utils", pattern = "\\.R$", full.names = TRUE)) {
    source(f)
}
```

## Read config and contrasts

```{r read-config}
config_data <- load_config(config_path)
cfg <- config_data$cfg
contrasts_df <- config_data$contrasts_df

cat("Config loaded:", config_path, "\n")
cat("Number of contrasts:", nrow(contrasts_df), "\n")
cat("Contrast IDs:", paste(contrasts_df$contrast_id, collapse = ", "), "\n")
```

## Build the contrast × stratum loop table

```{r build-loop}
loop_table <- do.call(rbind, lapply(seq_len(nrow(contrasts_df)), function(i) {
    row <- contrasts_df[i, ]
    strata <- parse_pipe_field(row$strata)
    data.frame(
        contrast_id = row$contrast_id,
        stratum     = strata,
        stringsAsFactors = FALSE
    )
}))

cat(sprintf("\nTotal jobs: %d contrasts × strata combinations\n", nrow(loop_table)))
print(loop_table)
```

## Pipeline settings summary

```{r settings-summary}
paired <- isTRUE(cfg$steps$paired_analysis)
cat("Paired analysis:", paired, "\n")
cat("RUVseq enabled:", isTRUE(cfg$steps$ruvseq), "\n")
cat("fGSEA enabled:", isTRUE(cfg$steps$fgsea), "\n")
cat("Results dir:", cfg$outputs$results_dir, "\n")
cat("Logs dir:", cfg$outputs$logs_dir, "\n")

if (paired) {
    cat("\n** Paired mode: steps 04 and 05 will be SKIPPED **\n")
    cat("** Pipeline flow: 01 → 02 → 03 → 06 → 07 **\n")
}
```
