---
title: "EasyDE Walkthrough — 00 Setup"
output: html_document
---

# Setup

This notebook loads all EasyDE utility functions, reads the pipeline config,
and defines the loop variables for the walkthrough.

**Run this first.** All subsequent notebooks depend on the variables set here.

## Set working directory

```{r set-root}
# EDIT THIS: path to your EasyDE pipeline root (where Snakefile lives)
pipeline_root <- normalizePath("../..")
knitr::opts_knit$set(root.dir = pipeline_root)
setwd(pipeline_root)
cat("Working directory:", getwd(), "\n")
```

## Choose your config

```{r config-choice}
# EDIT THIS: pick your config file
#   config/config_traits.yaml      — unpaired analysis
#   config/config_treatments.yaml  — paired analysis
config_path <- "config/config_treatments.yaml"
```

## Load libraries

```{r load-libs, message=FALSE, warning=FALSE}
suppressMessages({
    library(yaml)
    library(data.table)
    library(dplyr)
    library(tibble)
    library(stringr)
    library(ggplot2)
    library(DESeq2)
    library(RUVSeq)
    library(EDASeq)
    library(fgsea)
})
```

## Source all EasyDE utility functions

The pipeline ships a set of utility modules in `workflow/scripts/utils/`.
These are the "library" of reusable functions we'll use throughout:

- **`io_utils.R`** — `load_config()`, `load_count_matrix()`, `write_result()`, `parse_pipe_field()`, `make_summary_row()`
- **`filter_utils.R`** — `sanitize_names()`, `apply_subset_filters()`, sample/gene filtering
- **`validation_utils.R`** — `validate_design_matrix()`, `check_zero_inflation()`, `write_preflight_row()`
- **`stats_utils.R`** — `correlate_latent_factors()`, DESeq2 helpers
- **`plot_utils.R`** — `plot_volcano()`, `plot_fgsea_barplot()`
- **`logging_utils.R`** — structured logging

```{r source-utils}
for (f in list.files("workflow/scripts/utils", pattern = "\\.R$", full.names = TRUE)) {
    source(f)
}
cat("Loaded", length(list.files("workflow/scripts/utils", pattern = "\\.R$")),
    "utility modules\n")
```

## Read config and contrasts

```{r read-config}
config_data  <- load_config(config_path)
cfg          <- config_data$cfg
contrasts_df <- config_data$contrasts_df

cat("Config:", config_path, "\n")
cat("Contrasts:", nrow(contrasts_df), "\n\n")

# Show the contrasts table
contrasts_df[, c("contrast_id", "contrast_var", "control_grp",
                  "experimental_grp", "strata")]
```

## Pipeline mode

```{r pipeline-mode}
paired <- isTRUE(cfg$steps$paired_analysis)
cat("Paired analysis:", paired, "\n")
cat("RUVseq enabled:", isTRUE(cfg$steps$ruvseq), "\n")
cat("fGSEA enabled:", isTRUE(cfg$steps$fgsea), "\n")

if (paired) {
    cat("\n→ Paired mode: pipeline flow is 01 → 02 → 03 → 06 → 07\n")
    cat("  (Steps 04 and 05 are skipped — donor blocking replaces RUVSeq)\n")
}
```

## Build the contrast × stratum loop

```{r build-loop}
loop_table <- do.call(rbind, lapply(seq_len(nrow(contrasts_df)), function(i) {
    row <- contrasts_df[i, ]
    strata <- parse_pipe_field(row$strata)
    data.frame(
        contrast_id = row$contrast_id,
        stratum     = strata,
        stringsAsFactors = FALSE
    )
}))

cat(sprintf("Total: %d contrast × stratum combinations\n\n", nrow(loop_table)))
print(loop_table)
```
