---
title: "EasyDE Walkthrough — 02 Prepare Coldata"
output: html_document
---

# Step 02: Filter Samples & Build Coldata

For each contrast × stratum, this step:
- Loads and filters the count matrix
- Builds the coldata (sample metadata aligned to counts)
- Runs preflight quality checks (design rank, zero-inflation, etc.)
- Writes `coldata.csv`, `counts_filtered.csv`, and `preflight.csv`

```{r setup, include=FALSE}
pipeline_root <- normalizePath("../..")
knitr::opts_knit$set(root.dir = pipeline_root)
setwd(pipeline_root)
config_path <- "config/config_treatments.yaml"   # EDIT THIS
```

## Source the script

```{r source-script}
source("workflow/scripts/02_prepare_coldata.R")
```

## Build contrast × stratum loop

```{r build-loop}
config_data <- load_config(config_path)
contrasts_df <- config_data$contrasts_df

loop_table <- do.call(rbind, lapply(seq_len(nrow(contrasts_df)), function(i) {
    row <- contrasts_df[i, ]
    strata <- parse_pipe_field(row$strata)
    data.frame(contrast_id = row$contrast_id, stratum = strata,
               stringsAsFactors = FALSE)
}))
cat(sprintf("Processing %d contrast × stratum combinations\n", nrow(loop_table)))
```

## Run the loop

Each iteration filters samples and runs preflight checks for one stratum.
Strata that fail preflight are flagged but do **not** crash — downstream
steps will detect the preflight status and skip gracefully.

```{r run-loop}
for (i in seq_len(nrow(loop_table))) {
    cid <- loop_table$contrast_id[i]
    stm <- loop_table$stratum[i]
    cat(sprintf("\n=== [%d/%d] %s | %s ===\n", i, nrow(loop_table), cid, stm))

    tryCatch(
        main(config_path, cid, stm),
        error = function(e) cat(sprintf("  [ERROR] %s\n", e$message))
    )
}
```

## Check preflight results

```{r check-preflight}
for (i in seq_len(nrow(loop_table))) {
    cid <- loop_table$contrast_id[i]
    stm <- loop_table$stratum[i]
    cfg <- config_data$cfg
    inter_dir <- file.path(cfg$outputs$results_dir, cid, stm, "intermediates")
    status <- read_preflight_status(inter_dir)
    cat(sprintf("  %s | %s → preflight: %s\n", cid, stm, status))
}
```
