---
title: "EasyDE Walkthrough — 02 Prepare Coldata & Preflight QC"
output: html_document
---

# Step 02: Filter Samples, Build Coldata, and Preflight QC

For each contrast × stratum, this step:
1. Loads the raw pseudobulk count matrix
2. Filters samples (subset filters, group matching, donor pairing)
3. Filters genes (minimum count threshold)
4. Checks for zero-inflation
5. Builds the sanitized coldata for DESeq2
6. Validates the design matrix (rank, near-singularity)
7. Writes preflight report

```{r setup, include=FALSE}
pipeline_root <- normalizePath("../..")
knitr::opts_knit$set(root.dir = pipeline_root)
setwd(pipeline_root)

config_path <- "config/config_treatments.yaml"   # EDIT THIS

suppressMessages({
    library(yaml); library(data.table); library(dplyr)
    library(tibble); library(stringr)
})
for (f in list.files("workflow/scripts/utils", pattern = "\\.R$", full.names = TRUE)) source(f)
config_data <- load_config(config_path)
cfg <- config_data$cfg; contrasts_df <- config_data$contrasts_df
paired <- isTRUE(cfg$steps$paired_analysis)
```

## Process all strata

```{r process-loop}
for (row_i in seq_len(nrow(contrasts_df))) {

    contrast_row <- contrasts_df[row_i, ]
    contrast_id  <- contrast_row$contrast_id
    strata       <- parse_pipe_field(contrast_row$strata)

    for (stratum in strata) {
        stratum_safe <- sanitize_names(stratum)
        cat(sprintf("\n══════ %s | %s ══════\n", contrast_id, stratum))

        # --- Output directories ---
        out_dir <- file.path(cfg$outputs$results_dir, contrast_id,
                             stratum_safe, "intermediates")
        dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

        # --- Logger ---
        log_dir <- file.path(cfg$outputs$logs_dir, contrast_id)
        dir.create(log_dir, recursive = TRUE, showWarnings = FALSE)
        logger <- create_logger(
            log_dir, stratum_safe, "02",
            contrast_id = contrast_id, biosample = stratum
        )

        # ═══════════════════════════════════════════════
        # 1. Load count matrix
        # ═══════════════════════════════════════════════
        counts_path <- file.path(cfg$inputs$counts_dir,
                                 paste0(stratum_safe, ".counts.csv"))
        raw_mat <- load_count_matrix(counts_path)
        if (is.null(raw_mat)) {
            cat("  ✗ Count matrix not found — skipping\n"); next
        }

        gene_names <- raw_mat[[1]]
        raw_mat    <- raw_mat[, -1, drop = FALSE]
        rownames(raw_mat) <- gene_names
        cat(sprintf("  Loaded: %d genes × %d samples\n",
                    nrow(raw_mat), ncol(raw_mat)))

        # ═══════════════════════════════════════════════
        # 2. Load and filter metadata
        # ═══════════════════════════════════════════════
        sample_metadata <- fread(cfg$inputs$metadata_file) %>% as.data.frame()
        biosample_id_col <- contrast_row$biosample_id_col
        contrast_var     <- contrast_row$contrast_var
        control_grp      <- contrast_row$control_grp
        experimental_grp <- contrast_row$experimental_grp
        covariates       <- parse_pipe_field(contrast_row$covariates)
        strata_col       <- contrast_row$strata_col

        # Restrict metadata to samples present in the count matrix
        sample_metadata <- sample_metadata[
            sample_metadata[[biosample_id_col]] %in% colnames(raw_mat), ]

        # Apply subset filters (filter_col_1/filter_val_1, etc.)
        filters <- parse_contrast_filters(contrast_row)
        if (length(filters) > 0) {
            for (col_name in names(filters)) {
                allowed <- filters[[col_name]]
                sample_metadata <- sample_metadata[
                    sample_metadata[[col_name]] %in% allowed, ]
            }
            cat(sprintf("  After subset filters: %d samples\n",
                        nrow(sample_metadata)))
        }

        # Restrict to this stratum
        if (!is.na(strata_col) && strata_col != "") {
            sample_metadata <- sample_metadata[
                sample_metadata[[strata_col]] == stratum, ]
        }

        # Filter to control + experimental groups only
        is_categorical <- !is.na(control_grp) && control_grp != ""
        if (is_categorical) {
            sample_metadata <- sample_metadata[
                sample_metadata[[contrast_var]] %in%
                    c(control_grp, experimental_grp), ]
            cat(sprintf("  Group filter → %d samples\n",
                        nrow(sample_metadata)))
        }

        # Paired mode: keep only donors present in BOTH groups
        if (paired && is_categorical) {
            donor_col <- cfg$filtering$donor_col
            donors_ctrl <- unique(sample_metadata[
                sample_metadata[[contrast_var]] == control_grp, donor_col])
            donors_exp  <- unique(sample_metadata[
                sample_metadata[[contrast_var]] == experimental_grp, donor_col])
            paired_donors <- intersect(donors_ctrl, donors_exp)

            if (length(paired_donors) == 0) {
                cat("  ✗ No paired donors — skipping\n"); next
            }
            sample_metadata <- sample_metadata[
                sample_metadata[[donor_col]] %in% paired_donors, ]
            cat(sprintf("  Paired donors: %d (→ %d samples)\n",
                        length(paired_donors), nrow(sample_metadata)))
        }

        if (nrow(sample_metadata) == 0) {
            cat("  ✗ No samples remaining — skipping\n"); next
        }

        # Align count matrix to metadata
        keep_samples <- intersect(
            sample_metadata[[biosample_id_col]], colnames(raw_mat))
        raw_mat <- raw_mat[, keep_samples, drop = FALSE]
        rownames(sample_metadata) <- sample_metadata[[biosample_id_col]]
        sample_metadata <- sample_metadata[keep_samples, ]

        # ═══════════════════════════════════════════════
        # 3. Gene filtering
        # ═══════════════════════════════════════════════
        min_counts     <- cfg$filtering$min_gene_counts %||% 10
        genes_to_keep  <- rownames(raw_mat)[
            rowSums(raw_mat) >= min_counts]
        raw_mat <- raw_mat[genes_to_keep, , drop = FALSE]
        cat(sprintf("  Gene filter (≥%d total counts): %d genes retained\n",
                    min_counts, nrow(raw_mat)))

        # ═══════════════════════════════════════════════
        # 4. Zero-inflation check
        # ═══════════════════════════════════════════════
        zi_result <- check_zero_inflation(as.matrix(raw_mat))
        cat(sprintf("  Zero-inflation: %.1f%% of genes have ≥1 zero%s\n",
                    zi_result$pct_genes_with_zero,
                    if (zi_result$zero_inflated) " ⚠ REJECTED" else ""))

        # ═══════════════════════════════════════════════
        # 5. Prepare coldata (sanitize names, set factors)
        # ═══════════════════════════════════════════════
        coldata_result <- prepare_coldata(
            meta             = sample_metadata,
            biosample_id_col = biosample_id_col,
            contrast_var     = contrast_var,
            covariates       = covariates,
            scale_numeric    = isTRUE(cfg$deseq2$scale_numeric_covariates)
        )
        coldata      <- coldata_result$coldata
        name_map     <- coldata_result$name_map
        dropped_covs <- coldata_result$dropped_covariates

        if (length(dropped_covs) > 0) {
            cat(sprintf("  Dropped covariates (only 1 level): %s\n",
                        paste(dropped_covs, collapse = ", ")))
        }

        # ═══════════════════════════════════════════════
        # 6. Design matrix validation
        # ═══════════════════════════════════════════════
        safe_contrast <- sanitize_names(contrast_var)
        active_covs   <- setdiff(sanitize_names(covariates), dropped_covs)
        preflight_errors   <- character(0)
        preflight_warnings <- character(0)

        dm_result <- validate_design_matrix(
            contrast_var     = safe_contrast,
            covariates       = active_covs,
            meta             = coldata,
            control_grp      = if (is_categorical) control_grp else NULL,
            experimental_grp = if (is_categorical) experimental_grp else NULL
        )

        for (w in dm_result$warnings) preflight_warnings <- c(preflight_warnings, w)
        for (e in dm_result$errors)   preflight_errors   <- c(preflight_errors, e)
        if (zi_result$zero_inflated)  preflight_errors   <- c(preflight_errors,
            "100% of genes have ≥ 1 zero — data too sparse for DESeq2")

        preflight_status <- if (length(preflight_errors) > 0) "fail"
                            else if (length(preflight_warnings) > 0) "warn"
                            else "pass"

        cat(sprintf("  Preflight: %s (rank_ok=%s)\n",
                    toupper(preflight_status), dm_result$rank_ok))

        # ═══════════════════════════════════════════════
        # 7. Write outputs
        # ═══════════════════════════════════════════════
        write_result(coldata, file.path(out_dir, "coldata.csv"), sep = ",",
                     row_names = TRUE)
        fwrite(cbind(gene = rownames(raw_mat), as.data.frame(raw_mat)),
               file.path(out_dir, "counts_filtered.csv"))
        write_result(
            data.frame(original = names(name_map), sanitized = name_map,
                       stringsAsFactors = FALSE),
            file.path(out_dir, "name_map.csv"), sep = ","
        )
        write_preflight_row(
            row_data = list(
                contrast_id = contrast_id, stratum = stratum,
                status = preflight_status,
                n_genes = nrow(raw_mat), n_samples = nrow(coldata),
                n_control = sum(coldata[[safe_contrast]] == control_grp),
                n_experimental = sum(coldata[[safe_contrast]] == experimental_grp),
                n_covariates_active = length(active_covs),
                dropped_covariates = if (length(dropped_covs) > 0) paste(dropped_covs, collapse = "|") else NA,
                rank_ok = dm_result$rank_ok,
                near_singular_vars = NA,
                zero_inflated = zi_result$zero_inflated,
                pct_genes_with_zero = zi_result$pct_genes_with_zero,
                warnings = if (length(preflight_warnings) > 0) paste(preflight_warnings, collapse = "|") else NA,
                errors = if (length(preflight_errors) > 0) paste(preflight_errors, collapse = "|") else NA
            ),
            path = file.path(out_dir, "preflight.csv")
        )

        cat(sprintf("  ✓ Wrote coldata (%d samples), counts (%d genes), preflight\n",
                    nrow(coldata), nrow(raw_mat)))
    }
}
```

## Preflight summary

```{r preflight-summary}
cat("\n═══ Preflight Results ═══\n\n")
for (row_i in seq_len(nrow(contrasts_df))) {
    cid <- contrasts_df$contrast_id[row_i]
    strata <- parse_pipe_field(contrasts_df$strata[row_i])
    for (stm in strata) {
        inter_dir <- file.path(cfg$outputs$results_dir, cid,
                               sanitize_names(stm), "intermediates")
        status <- read_preflight_status(inter_dir)
        pf <- read_preflight_report(inter_dir)
        zi_flag <- if (!is.null(pf) && "zero_inflated" %in% colnames(pf))
                       ifelse(pf$zero_inflated[1], " [zero-inflated]", "") else ""
        cat(sprintf("  %s | %-20s → %s%s\n", cid, stm, status, zi_flag))
    }
}
```
