---
title: "EasyDE Walkthrough — 03 Base DESeq2"
output: html_document
---

# Step 03: Base DESeq2 Analysis

For each contrast × stratum that passed preflight, this step:
1. Loads the filtered coldata and count matrix from step 02
2. Builds the DESeq2 formula (`~ covariate1 + covariate2 + contrast_var`)
3. Creates a DESeqDataSet and runs DESeq2
4. Extracts results with apeglm shrinkage (if applicable)
5. Saves DE results and a volcano plot

In **paired mode**, the donor ID is included in the formula as a blocking
factor (e.g. `~ donor_accession + treatment`).

```{r setup, include=FALSE}
pipeline_root <- normalizePath("../..")
knitr::opts_knit$set(root.dir = pipeline_root)
setwd(pipeline_root)

config_path <- "config/config_treatments.yaml"   # EDIT THIS

suppressMessages({
    library(yaml); library(data.table); library(dplyr)
    library(DESeq2); library(ggplot2)
})
for (f in list.files("workflow/scripts/utils", pattern = "\\.R$", full.names = TRUE)) source(f)
config_data <- load_config(config_path)
cfg <- config_data$cfg; contrasts_df <- config_data$contrasts_df
paired <- isTRUE(cfg$steps$paired_analysis)
```

## Helper: build formula

The formula always ends with the contrast variable (DESeq2 convention).
In paired mode, the donor column is prepended as a blocking factor.

```{r formula-helper}
build_deseq_formula <- function(covariates, contrast_var) {
    terms <- c(covariates, contrast_var)
    paste0("~ ", paste(terms, collapse = " + "))
}
```

## Run DESeq2 for all strata

```{r run-deseq}
for (row_i in seq_len(nrow(contrasts_df))) {

    contrast_row <- contrasts_df[row_i, ]
    contrast_id  <- contrast_row$contrast_id
    strata       <- parse_pipe_field(contrast_row$strata)

    for (stratum in strata) {
        stratum_safe <- sanitize_names(stratum)
        inter_dir    <- file.path(cfg$outputs$results_dir, contrast_id,
                                  stratum_safe, "intermediates")
        finals_dir   <- file.path(cfg$outputs$results_dir, contrast_id,
                                  stratum_safe, "finals")
        plot_dir     <- file.path(cfg$outputs$results_dir, contrast_id,
                                  stratum_safe, "plots")
        dir.create(finals_dir, recursive = TRUE, showWarnings = FALSE)
        dir.create(plot_dir,   recursive = TRUE, showWarnings = FALSE)

        cat(sprintf("\n══════ %s | %s ══════\n", contrast_id, stratum))

        # --- Check preflight status ---
        preflight_status <- read_preflight_status(inter_dir)
        if (is.na(preflight_status) || preflight_status == "fail") {
            cat("  ✗ Preflight failed — skipping\n")
            writeLines("skipped=TRUE", file.path(finals_dir, "results_base.tsv"))
            next
        }

        # --- Load coldata and counts from step 02 ---
        coldata <- read.csv(file.path(inter_dir, "coldata.csv"),
                            check.names = FALSE, row.names = 1)
        counts_df <- fread(file.path(inter_dir, "counts_filtered.csv")) %>%
                     as.data.frame()
        gene_names <- counts_df[[1]]
        counts_mat <- as.matrix(counts_df[, -1])
        storage.mode(counts_mat) <- "integer"
        rownames(counts_mat) <- gene_names

        # --- Load name map for variable translations ---
        name_map_df <- read.csv(file.path(inter_dir, "name_map.csv"))
        name_map    <- setNames(name_map_df$sanitized, name_map_df$original)

        # --- Build formula ---
        contrast_var <- contrast_row$contrast_var
        safe_contrast <- sanitize_names(contrast_var)
        covariates_raw <- parse_pipe_field(contrast_row$covariates)
        safe_covs      <- sanitize_names(covariates_raw)
        safe_covs      <- safe_covs[safe_covs %in% colnames(coldata)]

        # In paired mode: add donor as a blocking factor
        if (paired) {
            donor_col <- sanitize_names(cfg$filtering$donor_col)
            if (donor_col %in% colnames(coldata) && !donor_col %in% safe_covs) {
                safe_covs <- c(donor_col, safe_covs)
            }
        }

        formula_str <- build_deseq_formula(safe_covs, safe_contrast)
        cat(sprintf("  Formula: %s\n", formula_str))

        # --- Align and create DESeqDataSet ---
        counts_mat <- counts_mat[, rownames(coldata), drop = FALSE]

        dds <- DESeqDataSetFromMatrix(
            countData = counts_mat,
            colData   = coldata,
            design    = as.formula(formula_str)
        )
        cat(sprintf("  DESeqDataSet: %d genes × %d samples\n",
                    nrow(dds), ncol(dds)))

        # --- Run DESeq2 ---
        dds <- DESeq(dds)

        # --- Extract results ---
        control_grp      <- contrast_row$control_grp
        experimental_grp <- contrast_row$experimental_grp
        is_categorical   <- !is.na(control_grp) && control_grp != ""
        lfc_shrinkage    <- isTRUE(cfg$deseq2$lfc_shrinkage)

        # Initialize logger for extract function
        log_dir <- file.path(cfg$outputs$logs_dir, contrast_id)
        dir.create(log_dir, recursive = TRUE, showWarnings = FALSE)
        logger <- create_logger(
            log_dir, stratum_safe, "03",
            contrast_id = contrast_id, biosample = stratum
        )

        deseq_out <- extract_deseq_results(
            dds              = dds,
            contrast_var     = safe_contrast,
            control_grp      = if (is_categorical) control_grp else NULL,
            experimental_grp = if (is_categorical) experimental_grp else NULL,
            lfc_shrinkage    = lfc_shrinkage,
            fdr_threshold    = cfg$deseq2$fdr_threshold,
            l2fc_threshold   = cfg$deseq2$l2fc_threshold,
            logger           = logger
        )

        cat(sprintf("  DEGs (FDR < %.2f, |LFC| > %.1f): %d\n",
                    cfg$deseq2$fdr_threshold, cfg$deseq2$l2fc_threshold,
                    deseq_out$n_degs))
        if (deseq_out$shrinkage_ran) cat("  ✓ apeglm shrinkage applied\n")

        # --- Write DE results ---
        write_result(deseq_out$results,
                     file.path(finals_dir, "results_base.tsv"))

        # Write model info
        model_info <- data.frame(formula = formula_str,
                                 n_degs = deseq_out$n_degs,
                                 stringsAsFactors = FALSE)
        write_result(model_info,
                     file.path(inter_dir, "model_info_base.csv"), sep = ",")

        # Save the dds object for step 04 (RUVseq needs it)
        saveRDS(dds, file.path(inter_dir, "dds_base.rds"))

        # --- Volcano plot ---
        plot_df <- deseq_out$results
        lfc_col <- if ("log2FoldChange_shrunk" %in% colnames(plot_df) &&
                       !all(is.na(plot_df$log2FoldChange_shrunk)))
                       "log2FoldChange_shrunk" else "log2FoldChange"

        p <- plot_volcano(
            df        = plot_df,
            lfc_col   = lfc_col,
            padj_col  = "padj",
            gene_col  = "gene",
            title     = sprintf("Volcano — %s %s (base)", stratum, contrast_id),
            fdr_line  = cfg$deseq2$fdr_threshold,
            l2fc_line = cfg$deseq2$l2fc_threshold
        )
        if (!is.null(p)) {
            pdf(file.path(plot_dir, "volcano_base.pdf"), width = 10, height = 8)
            print(p)
            dev.off()
        }

        cat(sprintf("  ✓ Wrote results_base.tsv, model_info, volcano\n"))
    }
}
```

## Quick summary: DEGs across strata

```{r deg-summary}
cat("\n═══ DEG Summary ═══\n\n")
for (row_i in seq_len(nrow(contrasts_df))) {
    cid <- contrasts_df$contrast_id[row_i]
    strata <- parse_pipe_field(contrasts_df$strata[row_i])
    for (stm in strata) {
        path <- file.path(cfg$outputs$results_dir, cid,
                          sanitize_names(stm), "finals", "results_base.tsv")
        if (file.exists(path) && !is_skip_sentinel(path)) {
            res <- read.delim(path)
            n <- sum(!is.na(res$padj) & res$padj < 0.05, na.rm = TRUE)
            cat(sprintf("  %s | %-20s → %d DEGs\n", cid, stm, n))
        } else {
            cat(sprintf("  %s | %-20s → skipped\n", cid, stm))
        }
    }
}
```
