---
title: "EasyDE Walkthrough — 03 Base DESeq2"
output: html_document
---

# Step 03: Base DESeq2 Analysis

For each contrast × stratum, fits the base DESeq2 model with user-specified
covariates. Produces `results_base.tsv` with log2 fold changes and adjusted
p-values.

In paired mode, the formula includes donor as a blocking factor
(e.g. `~ donor_accession + treatment`).

```{r setup, include=FALSE}
pipeline_root <- normalizePath("../..")
knitr::opts_knit$set(root.dir = pipeline_root)
setwd(pipeline_root)
config_path <- "config/config_treatments.yaml"   # EDIT THIS
```

## Source the script

```{r source-script}
source("workflow/scripts/03_run_deseq_base.R")
```

## Build loop

```{r build-loop}
config_data <- load_config(config_path)
contrasts_df <- config_data$contrasts_df

loop_table <- do.call(rbind, lapply(seq_len(nrow(contrasts_df)), function(i) {
    row <- contrasts_df[i, ]
    strata <- parse_pipe_field(row$strata)
    data.frame(contrast_id = row$contrast_id, stratum = strata,
               stringsAsFactors = FALSE)
}))
```

## Run the loop

Strata that were rejected at preflight (step 02) will be skipped
automatically — the script detects preflight status and exits cleanly.

```{r run-loop}
for (i in seq_len(nrow(loop_table))) {
    cid <- loop_table$contrast_id[i]
    stm <- loop_table$stratum[i]
    cat(sprintf("\n=== [%d/%d] %s | %s ===\n", i, nrow(loop_table), cid, stm))

    tryCatch(
        main(config_path, cid, stm),
        error = function(e) cat(sprintf("  [ERROR] %s\n", e$message))
    )
}
```

## Quick check: how many DEGs per stratum?

```{r check-degs}
cfg <- config_data$cfg
for (i in seq_len(nrow(loop_table))) {
    cid <- loop_table$contrast_id[i]
    stm <- loop_table$stratum[i]
    results_path <- file.path(cfg$outputs$results_dir, cid, stm, "finals", "results_base.tsv")
    if (file.exists(results_path) && !is_skip_sentinel(results_path)) {
        res <- read.delim(results_path)
        n_degs <- sum(!is.na(res$padj) & res$padj < 0.05, na.rm = TRUE)
        cat(sprintf("  %s | %s → %d DEGs (padj < 0.05)\n", cid, stm, n_degs))
    } else {
        cat(sprintf("  %s | %s → skipped\n", cid, stm))
    }
}
```
