---
title: "EasyDE Walkthrough — 04 RUVseq"
output: html_document
---

# Step 04: RUVseq Normalization

> **Paired mode:** This step is automatically **skipped** when
> `paired_analysis: true`. The donor blocking factor in the DESeq2 formula
> already accounts for inter-individual variation, making RUVseq redundant.

For unpaired analyses, this step uses empirical negative control genes
(genes with high p-values from step 03) to estimate latent factors of
unwanted variation via `RUVg`. It selects the optimal number of W factors
(k) using an F-statistic elbow method and prunes any W factors that
correlate with the contrast variable.

```{r setup, include=FALSE}
pipeline_root <- normalizePath("../..")
knitr::opts_knit$set(root.dir = pipeline_root)
setwd(pipeline_root)
config_path <- "config/config_treatments.yaml"   # EDIT THIS
```

## Check if RUVseq should run

```{r check-mode}
cfg <- suppressWarnings(yaml::read_yaml(config_path))
paired <- isTRUE(cfg$steps$paired_analysis)

if (paired) {
    cat("** PAIRED MODE — this entire notebook is SKIPPED **\n")
    cat("** Donor is already a blocking factor in the DESeq2 formula **\n")
    knitr::knit_exit()
}
```

## Source the script and run

```{r source-script}
source("workflow/scripts/04_run_ruvseq.R")
```

```{r build-loop}
config_data <- load_config(config_path)
contrasts_df <- config_data$contrasts_df

loop_table <- do.call(rbind, lapply(seq_len(nrow(contrasts_df)), function(i) {
    row <- contrasts_df[i, ]
    strata <- parse_pipe_field(row$strata)
    data.frame(contrast_id = row$contrast_id, stratum = strata,
               stringsAsFactors = FALSE)
}))
```

```{r run-loop}
for (i in seq_len(nrow(loop_table))) {
    cid <- loop_table$contrast_id[i]
    stm <- loop_table$stratum[i]
    cat(sprintf("\n=== [%d/%d] %s | %s ===\n", i, nrow(loop_table), cid, stm))

    tryCatch(
        main(config_path, cid, stm),
        error = function(e) cat(sprintf("  [ERROR] %s\n", e$message))
    )
}
```

## Check optimal k per stratum

```{r check-k}
for (i in seq_len(nrow(loop_table))) {
    cid <- loop_table$contrast_id[i]
    stm <- loop_table$stratum[i]
    summary_path <- file.path(cfg$outputs$results_dir, cid, stm,
                              "intermediates", "ruvseq_summary.csv")
    if (file.exists(summary_path) && !is_skip_sentinel(summary_path)) {
        summary <- read.csv(summary_path)
        cat(sprintf("  %s | %s → best k = %s\n", cid, stm,
                    summary$best_k[1]))
    } else {
        cat(sprintf("  %s | %s → skipped\n", cid, stm))
    }
}
```
