---
title: "EasyDE Walkthrough — 04 RUVseq Normalization"
output: html_document
---

# Step 04: RUVseq — Remove Unwanted Variation

> **Paired mode:** This entire step is **skipped**. The donor blocking factor
> in the DESeq2 formula already captures inter-individual variation.

For unpaired analyses, RUVseq identifies and removes technical batch effects
using empirical negative control genes. Here's how:

1. **Identify negative controls** — genes with high p-values (p > 0.5) from
   the base DESeq2 results (step 03). These genes are not DE, so their
   variation is "unwanted" technical noise.
2. **Run RUVg** for k = 1 through max_k — each k adds another latent W factor.
3. **Select optimal k** — compute an RLE ANOVA F-statistic for each k.
   The "elbow" of the curve (maximum second derivative) is the optimal k.
4. **Prune W factors** — test each W factor for correlation with the
   contrast variable. Correlated Ws are removed to preserve real biology.

```{r setup, include=FALSE}
pipeline_root <- normalizePath("../..")
knitr::opts_knit$set(root.dir = pipeline_root)
setwd(pipeline_root)

config_path <- "config/config_treatments.yaml"   # EDIT THIS

suppressMessages({
    library(yaml); library(data.table); library(dplyr)
    library(tibble); library(tidyr)
    library(RUVSeq); library(EDASeq)
})
for (f in list.files("workflow/scripts/utils", pattern = "\\.R$", full.names = TRUE)) source(f)
config_data <- load_config(config_path)
cfg <- config_data$cfg; contrasts_df <- config_data$contrasts_df
```

## Check paired mode

```{r check-paired}
paired <- isTRUE(cfg$steps$paired_analysis)

if (paired) {
    cat("╔══════════════════════════════════════════════╗\n")
    cat("║  PAIRED MODE — RUVseq is SKIPPED            ║\n")
    cat("║  Donor blocking already handles batch.       ║\n")
    cat("╚══════════════════════════════════════════════╝\n")
    knitr::knit_exit()
}
```

## Run RUVseq for all strata

```{r run-ruvseq}
for (row_i in seq_len(nrow(contrasts_df))) {

    contrast_row <- contrasts_df[row_i, ]
    contrast_id  <- contrast_row$contrast_id
    strata       <- parse_pipe_field(contrast_row$strata)

    for (stratum in strata) {
        stratum_safe <- sanitize_names(stratum)
        inter_dir    <- file.path(cfg$outputs$results_dir, contrast_id,
                                  stratum_safe, "intermediates")
        finals_dir   <- file.path(cfg$outputs$results_dir, contrast_id,
                                  stratum_safe, "finals")

        cat(sprintf("\n══════ %s | %s ══════\n", contrast_id, stratum))

        # --- Check preflight ---
        preflight_status <- read_preflight_status(inter_dir)
        if (is.na(preflight_status) || preflight_status == "fail") {
            cat("  ✗ Preflight failed — skipping\n"); next
        }

        # --- Load counts and base results ---
        counts_df  <- fread(file.path(inter_dir, "counts_filtered.csv")) %>%
                      as.data.frame()
        gene_names <- counts_df[[1]]
        counts_mat <- as.matrix(counts_df[, -1])
        storage.mode(counts_mat) <- "integer"
        rownames(counts_mat) <- gene_names

        base_results <- fread(file.path(finals_dir, "results_base.tsv")) %>%
                        as.data.frame()
        if (is_skip_sentinel(file.path(finals_dir, "results_base.tsv"))) {
            cat("  ✗ Base DESeq2 was skipped — skipping RUVseq\n"); next
        }

        coldata <- read.csv(file.path(inter_dir, "coldata.csv"),
                            check.names = FALSE, row.names = 1)
        biosample_id_col <- sanitize_names(contrast_row$biosample_id_col)

        # --- Identify covariates ---
        safe_contrast <- sanitize_names(contrast_row$contrast_var)
        safe_covs     <- sanitize_names(parse_pipe_field(contrast_row$covariates))
        safe_covs     <- safe_covs[safe_covs %in% colnames(coldata)]

        # --- Determine max k ---
        n_samples <- ncol(counts_mat)
        n_covs    <- length(safe_covs)
        max_k_cfg <- cfg$ruvseq$max_latent_factors %||% 10
        max_k     <- min(max_k_cfg, floor(n_samples / 2) - n_covs - 1)
        max_k     <- max(max_k, 1)

        cat(sprintf("  max_k = %d (samples=%d, covs=%d)\n",
                    max_k, n_samples, n_covs))

        # ═══════════════════════════════════════════════
        # 1. Run RUVg for all k values
        # ═══════════════════════════════════════════════
        empirical_thresh <- cfg$ruvseq$empirical_pval_threshold %||% 0.5

        ruvseq_results <- run_ruvseq_all_k(
            counts_mat            = counts_mat,
            base_results          = base_results,
            max_k                 = max_k,
            empirical_pval_thresh = empirical_thresh
        )
        cat(sprintf("  Negative controls (p > %.1f): %d genes\n",
                    empirical_thresh,
                    sum(!is.na(base_results$pvalue) &
                        base_results$pvalue > empirical_thresh)))

        # ═══════════════════════════════════════════════
        # 2. F-statistic elbow: find optimal k
        # ═══════════════════════════════════════════════
        anova_df <- compute_rle_anova(
            ruvseq_results   = ruvseq_results,
            biosample_id_col = biosample_id_col,
            contrast_id      = contrast_id,
            stratum          = stratum
        )

        # Elbow detection: maximum second derivative
        f_vals   <- anova_df$f_statistic
        second_d <- diff(diff(f_vals))
        best_k   <- which.max(second_d) + 1
        best_k   <- max(1, min(best_k, max_k))

        cat(sprintf("  F-statistic elbow → best_k = %d\n", best_k))

        # ═══════════════════════════════════════════════
        # 3. Prune W factors correlated with contrast
        # ═══════════════════════════════════════════════
        W_best <- ruvseq_results[[best_k]]$W
        latent_names <- paste0("W_", 1:best_k)
        colnames(W_best) <- latent_names

        coldata_W <- cbind(coldata, W_best[rownames(coldata), , drop = FALSE])

        latent_corr_raw <- parse_pipe_field(contrast_row$latent_corr_vars)
        latent_corr_vars <- sanitize_names(latent_corr_raw)
        latent_corr_vars <- latent_corr_vars[latent_corr_vars %in% colnames(coldata)]

        if (length(latent_corr_vars) > 0) {
            ws_result <- identify_disease_ws(
                coldata_with_W   = coldata_W,
                latent_names     = latent_names,
                latent_corr_vars = c(latent_corr_vars, safe_contrast),
                contrast_var     = safe_contrast,
                biosample_id_col = biosample_id_col
            )
            disease_ws <- ws_result$disease_associated_Ws
            safe_ws    <- ws_result$safe_Ws
        } else {
            disease_ws <- character(0)
            safe_ws    <- latent_names
        }

        if (length(disease_ws) > 0) {
            cat(sprintf("  ⚠ Pruned (correlated with %s): %s\n",
                        safe_contrast, paste(disease_ws, collapse = ", ")))
        }
        cat(sprintf("  Safe W factors: %s\n",
                    if (length(safe_ws) > 0) paste(safe_ws, collapse = ", ") else "none"))

        # ═══════════════════════════════════════════════
        # 4. Write outputs
        # ═══════════════════════════════════════════════
        write_result(coldata_W,
                     file.path(inter_dir, "ruvseq_best_coldata.csv"),
                     sep = ",", row_names = TRUE)
        write_result(anova_df,
                     file.path(inter_dir, "ruvseq_anova.csv"), sep = ",")

        ruvseq_summary <- data.frame(
            contrast_id = contrast_id,
            stratum     = stratum,
            best_k      = best_k,
            max_k_tested = max_k,
            safe_ws     = paste(safe_ws, collapse = "|"),
            pruned_ws   = paste(disease_ws, collapse = "|"),
            empirical_pval_threshold = empirical_thresh,
            stringsAsFactors = FALSE
        )
        write_result(ruvseq_summary,
                     file.path(inter_dir, "ruvseq_summary.csv"), sep = ",")

        cat("  ✓ Wrote ruvseq_best_coldata, ruvseq_anova, ruvseq_summary\n")
    }
}
```
