---
title: "EasyDE Walkthrough — 05 Final DESeq2 with W Factors"
output: html_document
---

# Step 05: Final DESeq2 (RUVseq-corrected)

> **Paired mode:** This step is **skipped** — the base model from step 03
> is the final model.

For unpaired analyses, this step re-fits DESeq2 with the safe W factors
from step 04 added to the formula. This removes unwanted technical variation
while preserving genuine biological signal.

**Formula:** `~ W_1 + W_2 + ... + covariate + contrast_var`
(only safe Ws are included — correlated ones were pruned in step 04)

```{r setup, include=FALSE}
pipeline_root <- normalizePath("../..")
knitr::opts_knit$set(root.dir = pipeline_root)
setwd(pipeline_root)

config_path <- "config/config_treatments.yaml"   # EDIT THIS

suppressMessages({
    library(yaml); library(data.table); library(dplyr)
    library(DESeq2); library(ggplot2)
})
for (f in list.files("workflow/scripts/utils", pattern = "\\.R$", full.names = TRUE)) source(f)
config_data <- load_config(config_path)
cfg <- config_data$cfg; contrasts_df <- config_data$contrasts_df
```

## Check paired mode

```{r check-paired}
paired <- isTRUE(cfg$steps$paired_analysis)

if (paired) {
    cat("╔══════════════════════════════════════════════╗\n")
    cat("║  PAIRED MODE — this step is SKIPPED          ║\n")
    cat("║  Step 03 base model IS the final model.      ║\n")
    cat("╚══════════════════════════════════════════════╝\n")
    knitr::knit_exit()
}
```

## Re-fit DESeq2 with W factors

```{r run-deseq-ruv}
build_deseq_formula <- function(covariates, contrast_var) {
    terms <- c(covariates, contrast_var)
    paste0("~ ", paste(terms, collapse = " + "))
}

for (row_i in seq_len(nrow(contrasts_df))) {

    contrast_row <- contrasts_df[row_i, ]
    contrast_id  <- contrast_row$contrast_id
    strata       <- parse_pipe_field(contrast_row$strata)

    for (stratum in strata) {
        stratum_safe <- sanitize_names(stratum)
        inter_dir    <- file.path(cfg$outputs$results_dir, contrast_id,
                                  stratum_safe, "intermediates")
        finals_dir   <- file.path(cfg$outputs$results_dir, contrast_id,
                                  stratum_safe, "finals")
        plot_dir     <- file.path(cfg$outputs$results_dir, contrast_id,
                                  stratum_safe, "plots")
        dir.create(finals_dir, recursive = TRUE, showWarnings = FALSE)
        dir.create(plot_dir,   recursive = TRUE, showWarnings = FALSE)

        cat(sprintf("\n══════ %s | %s ══════\n", contrast_id, stratum))

        # --- Check preflight + RUVseq summary ---
        ruvseq_summary_path <- file.path(inter_dir, "ruvseq_summary.csv")
        if (!file.exists(ruvseq_summary_path) ||
            is_skip_sentinel(ruvseq_summary_path)) {
            cat("  ✗ RUVseq was skipped — writing sentinel\n")
            writeLines("skipped=TRUE", file.path(finals_dir, "results_ruv.tsv"))
            next
        }

        # --- Read RUVseq outputs ---
        ruvseq_summary <- read.csv(ruvseq_summary_path)
        safe_ws <- strsplit(ruvseq_summary$safe_ws[1], "\\|")[[1]]
        safe_ws <- safe_ws[safe_ws != ""]

        if (length(safe_ws) == 0) {
            cat("  ⚠ All W factors were pruned — RUV model = base model\n")
            writeLines("skipped=TRUE", file.path(finals_dir, "results_ruv.tsv"))
            next
        }

        # --- Load coldata with W factors + counts ---
        coldata_W <- read.csv(file.path(inter_dir, "ruvseq_best_coldata.csv"),
                              check.names = FALSE, row.names = 1)
        counts_df <- fread(file.path(inter_dir, "counts_filtered.csv")) %>%
                     as.data.frame()
        gene_names <- counts_df[[1]]
        counts_mat <- as.matrix(counts_df[, -1])
        storage.mode(counts_mat) <- "integer"
        rownames(counts_mat) <- gene_names

        # --- Build formula with safe Ws ---
        safe_contrast <- sanitize_names(contrast_row$contrast_var)
        safe_covs     <- sanitize_names(parse_pipe_field(contrast_row$covariates))
        safe_covs     <- safe_covs[safe_covs %in% colnames(coldata_W)]

        all_terms    <- c(safe_ws, safe_covs)
        formula_str  <- build_deseq_formula(all_terms, safe_contrast)
        cat(sprintf("  Formula: %s\n", formula_str))

        # --- Create DESeqDataSet and run ---
        counts_mat <- counts_mat[, rownames(coldata_W), drop = FALSE]

        dds <- DESeqDataSetFromMatrix(
            countData = counts_mat,
            colData   = coldata_W,
            design    = as.formula(formula_str)
        )
        dds <- DESeq(dds)

        # --- Extract results ---
        control_grp      <- contrast_row$control_grp
        experimental_grp <- contrast_row$experimental_grp
        is_categorical   <- !is.na(control_grp) && control_grp != ""
        lfc_shrinkage    <- isTRUE(cfg$deseq2$lfc_shrinkage)

        log_dir <- file.path(cfg$outputs$logs_dir, contrast_id)
        logger <- create_logger(
            log_dir, stratum_safe, "05",
            contrast_id = contrast_id, biosample = stratum
        )

        deseq_out <- extract_deseq_results(
            dds              = dds,
            contrast_var     = safe_contrast,
            control_grp      = if (is_categorical) control_grp else NULL,
            experimental_grp = if (is_categorical) experimental_grp else NULL,
            lfc_shrinkage    = lfc_shrinkage,
            fdr_threshold    = cfg$deseq2$fdr_threshold,
            l2fc_threshold   = cfg$deseq2$l2fc_threshold,
            logger           = logger
        )

        cat(sprintf("  DEGs (RUV-corrected): %d\n", deseq_out$n_degs))

        # --- Write outputs ---
        write_result(deseq_out$results,
                     file.path(finals_dir, "results_ruv.tsv"))

        model_info <- data.frame(formula = formula_str,
                                 n_degs = deseq_out$n_degs,
                                 stringsAsFactors = FALSE)
        write_result(model_info,
                     file.path(inter_dir, "model_info_ruv.csv"), sep = ",")

        # Volcano
        plot_df <- deseq_out$results
        lfc_col <- if ("log2FoldChange_shrunk" %in% colnames(plot_df) &&
                       !all(is.na(plot_df$log2FoldChange_shrunk)))
                       "log2FoldChange_shrunk" else "log2FoldChange"

        p <- plot_volcano(
            df = plot_df, lfc_col = lfc_col, padj_col = "padj",
            gene_col = "gene",
            title = sprintf("Volcano — %s %s (RUV)", stratum, contrast_id),
            fdr_line = cfg$deseq2$fdr_threshold,
            l2fc_line = cfg$deseq2$l2fc_threshold
        )
        if (!is.null(p)) {
            pdf(file.path(plot_dir, "volcano_ruv.pdf"), width = 10, height = 8)
            print(p); dev.off()
        }

        cat("  ✓ Wrote results_ruv.tsv, model_info_ruv, volcano_ruv\n")
    }
}
```
