---
title: "EasyDE Walkthrough — 06 fGSEA"
output: html_document
---

# Step 06: Gene Set Enrichment Analysis (fGSEA)

For each contrast × stratum, this step:
1. **Loads DE results** — uses the Wald test statistic as the ranking metric
2. **Excludes housekeeping genes** — ribosomal (RPL/RPS) and mitochondrial
   (MT-) genes are removed before ranking to prevent them from dominating
3. **Ranks genes** — sorted by the chosen statistic (default: Wald stat)
4. **Runs fGSEA** — computes enrichment scores for Reactome + KEGG pathways
5. **Saves results and barplot**

```{r setup, include=FALSE}
pipeline_root <- normalizePath("../..")
knitr::opts_knit$set(root.dir = pipeline_root)
setwd(pipeline_root)

config_path <- "config/config_treatments.yaml"   # EDIT THIS

suppressMessages({
    library(yaml); library(data.table); library(dplyr)
    library(fgsea); library(ggplot2)
})
for (f in list.files("workflow/scripts/utils", pattern = "\\.R$", full.names = TRUE)) source(f)
config_data <- load_config(config_path)
cfg <- config_data$cfg; contrasts_df <- config_data$contrasts_df
paired <- isTRUE(cfg$steps$paired_analysis)
```

## Load gene set pathways

```{r load-pathways}
gmt_file <- cfg$fgsea$gene_sets_file
pathways <- fgsea::gmtPathways(gmt_file)
cat(sprintf("Loaded %d pathways from %s\n", length(pathways), basename(gmt_file)))
```

## Build gene exclusion list

Ribosomal and mitochondrial genes are excluded because their high and
variable expression dominates rankings without being biologically
informative for most analyses.

```{r exclusion-list}
# Load exclusion gene lists
exclude_files <- cfg$fgsea$exclude_gene_lists
if (is.null(exclude_files)) exclude_files <- list()

# Load pattern-based exclusions (e.g. "^MT-")
exclude_patterns <- cfg$fgsea$exclude_patterns
if (is.null(exclude_patterns)) exclude_patterns <- character(0)

# We'll build the full list per-stratum since gene names vary
cat(sprintf("Exclusion: %d gene list files + %d patterns\n",
            length(exclude_files), length(exclude_patterns)))
```

## Run fGSEA for all strata

```{r run-fgsea}
# Which DE results to run fGSEA on
run_on <- cfg$fgsea$run_on %||% "both"
if (paired) run_on <- "base"   # paired mode: only base results exist

suffixes <- switch(run_on,
    "base" = "base",
    "ruv"  = "ruv",
    "both" = c("base", "ruv"),
    "base"   # fallback
)
cat(sprintf("Running fGSEA on: %s\n", paste(suffixes, collapse = ", ")))

for (row_i in seq_len(nrow(contrasts_df))) {

    contrast_row <- contrasts_df[row_i, ]
    contrast_id  <- contrast_row$contrast_id
    strata       <- parse_pipe_field(contrast_row$strata)

    for (stratum in strata) {
        stratum_safe <- sanitize_names(stratum)
        inter_dir    <- file.path(cfg$outputs$results_dir, contrast_id,
                                  stratum_safe, "intermediates")
        finals_dir   <- file.path(cfg$outputs$results_dir, contrast_id,
                                  stratum_safe, "finals")
        plot_dir     <- file.path(cfg$outputs$results_dir, contrast_id,
                                  stratum_safe, "plots")
        dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)

        log_dir <- file.path(cfg$outputs$logs_dir, contrast_id)
        dir.create(log_dir, recursive = TRUE, showWarnings = FALSE)
        logger <- create_logger(
            log_dir, stratum_safe, "06",
            contrast_id = contrast_id, biosample = stratum
        )

        cat(sprintf("\n══════ %s | %s ══════\n", contrast_id, stratum))

        for (suffix in suffixes) {
            results_path <- file.path(finals_dir,
                                       sprintf("results_%s.tsv", suffix))
            if (!file.exists(results_path) || is_skip_sentinel(results_path)) {
                cat(sprintf("  ✗ results_%s.tsv not found — skipping\n", suffix))
                writeLines("skipped=TRUE",
                           file.path(finals_dir,
                                     sprintf("fgsea_%s.tsv", suffix)))
                next
            }

            # --- Load DE results ---
            de_results <- read.delim(results_path)
            cat(sprintf("  [%s] %d genes in DE results\n",
                        suffix, nrow(de_results)))

            # --- Build exclusion list ---
            static_excluded <- build_exclusion_list(
                exclude_gene_lists = exclude_files,
                exclude_patterns   = exclude_patterns,
                all_genes          = de_results$gene
            )
            cat(sprintf("  [%s] Excluding %d genes (ribo/mito)\n",
                        suffix, length(static_excluded)))

            # Combine with statistically-driven exclusions
            all_excluded <- unique(c(
                static_excluded,
                de_results$gene[is.na(de_results$stat)]
            ))

            # --- Rank genes ---
            ranking_stat <- cfg$fgsea$ranking_stat %||% "stat"
            ranks <- build_gene_ranks(
                deseq_results = de_results,
                exclude_genes = all_excluded,
                ranking_stat  = ranking_stat
            )
            cat(sprintf("  [%s] Ranked %d genes by '%s'\n",
                        suffix, length(ranks), ranking_stat))

            if (length(ranks) == 0) {
                cat(sprintf("  [%s] No genes to rank — skipping\n", suffix))
                next
            }

            # --- Run fGSEA ---
            min_size <- cfg$fgsea$min_gene_set_size %||% 10
            max_size <- cfg$fgsea$max_gene_set_size %||% 500
            fdr_threshold <- cfg$fgsea$fdr_threshold %||% 0.10

            result <- run_fgsea_analysis(
                ranks         = ranks,
                pathways      = pathways,
                min_size      = min_size,
                max_size      = max_size,
                fdr_threshold = fdr_threshold,
                logger        = logger,
                label         = sprintf("fgsea_%s", suffix)
            )

            cat(sprintf("  [%s] %d significant pathways (FDR < %.2f)\n",
                        suffix, result$n_sig, fdr_threshold))

            # --- Write results ---
            write_result(result$all,
                         file.path(finals_dir,
                                   sprintf("fgsea_%s.tsv", suffix)))

            # --- Barplot ---
            p <- plot_fgsea_barplot(
                fgsea_sig_df = result$sig,
                top_n        = cfg$fgsea$n_top_pathways_plot %||% 20,
                title        = sprintf("Top pathways — %s %s (%s)",
                                       stratum, contrast_id, suffix)
            )
            if (!is.null(p)) {
                pdf(file.path(plot_dir, sprintf("fgsea_%s.pdf", suffix)),
                    width = 12, height = 8)
                print(p); dev.off()
            }
        }
        cat("  ✓ fGSEA complete\n")
    }
}
```

## Pathway summary

```{r pathway-summary}
cat("\n═══ Pathway Summary ═══\n\n")
for (row_i in seq_len(nrow(contrasts_df))) {
    cid <- contrasts_df$contrast_id[row_i]
    strata <- parse_pipe_field(contrasts_df$strata[row_i])
    for (stm in strata) {
        for (sfx in suffixes) {
            path <- file.path(cfg$outputs$results_dir, cid,
                              sanitize_names(stm), "finals",
                              sprintf("fgsea_%s.tsv", sfx))
            if (file.exists(path) && !is_skip_sentinel(path)) {
                res <- read.delim(path)
                n_sig <- sum(!is.na(res$padj) & res$padj < 0.05, na.rm = TRUE)
                cat(sprintf("  %s | %-20s [%s] → %d significant pathways\n",
                            cid, stm, sfx, n_sig))
            }
        }
    }
}
```
