---
title: "EasyDE Walkthrough ŌĆö 07 Aggregate Results"
output: html_document
---

# Step 07: Aggregate Results & Status Overview

Runs **once per contrast**. Collects all per-stratum outputs and produces:
- `contrast_summary.csv` ŌĆö one row per stratum with DEG counts, pathway
  counts, and pipeline status
- `preflight_summary.csv` ŌĆö preflight QC across all strata
- `status_tile.pdf` ŌĆö heatmap showing which strata succeeded or were skipped
- Merged DE and fGSEA tables across all strata

```{r setup, include=FALSE}
pipeline_root <- normalizePath("../..")
knitr::opts_knit$set(root.dir = pipeline_root)
setwd(pipeline_root)

config_path <- "config/config_treatments.yaml"   # EDIT THIS

suppressMessages({
    library(yaml); library(data.table); library(dplyr); library(ggplot2)
})
for (f in list.files("workflow/scripts/utils", pattern = "\\.R$", full.names = TRUE)) source(f)
config_data <- load_config(config_path)
cfg <- config_data$cfg; contrasts_df <- config_data$contrasts_df
paired <- isTRUE(cfg$steps$paired_analysis)
```

## Aggregate each contrast

```{r aggregate-contrasts}
for (row_i in seq_len(nrow(contrasts_df))) {

    contrast_row <- contrasts_df[row_i, ]
    contrast_id  <- contrast_row$contrast_id
    strata       <- parse_pipe_field(contrast_row$strata)
    summary_dir  <- file.path(cfg$outputs$results_dir, contrast_id, "summary")
    dir.create(summary_dir, recursive = TRUE, showWarnings = FALSE)

    cat(sprintf("\nŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉ Aggregating: %s ŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉ\n", contrast_id))

    # ŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉ
    # 1. Collect per-stratum summaries
    # ŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉ
    summary_rows <- list()
    all_de_base  <- list()
    all_de_ruv   <- list()
    all_fgsea_base <- list()
    all_fgsea_ruv  <- list()
    preflight_rows <- list()

    for (stratum in strata) {
        stratum_safe <- sanitize_names(stratum)
        inter_dir    <- file.path(cfg$outputs$results_dir, contrast_id,
                                  stratum_safe, "intermediates")
        finals_dir   <- file.path(cfg$outputs$results_dir, contrast_id,
                                  stratum_safe, "finals")

        # Determine status
        preflight_status <- read_preflight_status(inter_dir)

        if (is.na(preflight_status)) {
            # Coldata was never produced ŌåÆ filtering removed all samples
            status <- "skipped_filtering"
        } else if (preflight_status == "fail") {
            status <- "skipped_preflight"
        } else {
            # Preflight passed ŌĆö check if DESeq2 results exist
            base_path <- file.path(finals_dir, "results_base.tsv")
            if (!file.exists(base_path) || is_skip_sentinel(base_path)) {
                status <- "error"
            } else {
                status <- "success"
            }
        }

        cat(sprintf("  %-20s ŌåÆ %s\n", stratum, status))

        # Count DEGs and pathways
        n_degs_base <- n_degs_ruv <- n_pathways_base <- n_pathways_ruv <- NA

        for (suffix in c("base", "ruv")) {
            de_path <- file.path(finals_dir, sprintf("results_%s.tsv", suffix))
            if (file.exists(de_path) && !is_skip_sentinel(de_path)) {
                de <- read.delim(de_path)
                n_degs <- sum(!is.na(de$padj) & de$padj < 0.05, na.rm = TRUE)
                de$biosample <- stratum

                if (suffix == "base") {
                    n_degs_base <- n_degs
                    all_de_base[[stratum]] <- de
                } else {
                    n_degs_ruv <- n_degs
                    all_de_ruv[[stratum]] <- de
                }
            }

            fgsea_path <- file.path(finals_dir, sprintf("fgsea_%s.tsv", suffix))
            if (file.exists(fgsea_path) && !is_skip_sentinel(fgsea_path)) {
                fg <- read.delim(fgsea_path)
                n_sig <- sum(!is.na(fg$padj) & fg$padj < 0.05, na.rm = TRUE)
                fg$biosample <- stratum

                if (suffix == "base") {
                    n_pathways_base <- n_sig
                    all_fgsea_base[[stratum]] <- fg
                } else {
                    n_pathways_ruv <- n_sig
                    all_fgsea_ruv[[stratum]] <- fg
                }
            }
        }

        # Collect preflight report
        pf_report <- read_preflight_report(inter_dir)
        if (!is.null(pf_report)) preflight_rows[[stratum]] <- pf_report

        # Summary row
        summary_rows[[stratum]] <- data.frame(
            contrast = contrast_id,
            biosample = stratum,
            status = status,
            n_degs_base = n_degs_base,
            n_degs_ruv = n_degs_ruv,
            n_pathways_base = n_pathways_base,
            n_pathways_ruv = n_pathways_ruv,
            stringsAsFactors = FALSE
        )
    }

    # ŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉ
    # 2. Write summary files
    # ŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉŌĢÉ
    contrast_summary <- do.call(rbind, summary_rows)
    write.csv(contrast_summary,
              file.path(summary_dir, "contrast_summary.csv"),
              row.names = FALSE)

    if (length(preflight_rows) > 0) {
        pf_summary <- do.call(rbind, preflight_rows)
        write.csv(pf_summary,
                  file.path(summary_dir, "preflight_summary.csv"),
                  row.names = FALSE)
    }

    # Merged DE tables
    if (length(all_de_base) > 0)
        fwrite(do.call(rbind, all_de_base),
               file.path(summary_dir,
                         sprintf("%s_results_base.tsv", contrast_id)),
               sep = "\t")
    if (length(all_de_ruv) > 0)
        fwrite(do.call(rbind, all_de_ruv),
               file.path(summary_dir,
                         sprintf("%s_results_ruv.tsv", contrast_id)),
               sep = "\t")

    # Merged fGSEA tables
    if (length(all_fgsea_base) > 0)
        fwrite(do.call(rbind, all_fgsea_base),
               file.path(summary_dir,
                         sprintf("%s_fgsea_base.tsv", contrast_id)),
               sep = "\t")
    if (length(all_fgsea_ruv) > 0)
        fwrite(do.call(rbind, all_fgsea_ruv),
               file.path(summary_dir,
                         sprintf("%s_fgsea_ruv.tsv", contrast_id)),
               sep = "\t")

    cat(sprintf("  Ō£ō Wrote summary to %s\n", summary_dir))
}
```

## Status tile plot (all contrasts)

This heatmap shows the pipeline status for every stratum ├Ś contrast
combination. The numbers overlaid are the base DEG counts.

```{r status-tile, fig.width=14, fig.height=8}
all_summaries <- list()
for (row_i in seq_len(nrow(contrasts_df))) {
    cid <- contrasts_df$contrast_id[row_i]
    path <- file.path(cfg$outputs$results_dir, cid,
                      "summary", "contrast_summary.csv")
    if (file.exists(path))
        all_summaries[[cid]] <- read.csv(path, check.names = FALSE)
}

if (length(all_summaries) > 0) {
    combined <- do.call(rbind, all_summaries)

    # Status levels and colors
    status_levels <- c("success", "error", "skipped_preflight",
                       "skipped_filtering", "not_run")
    status_colors <- c(
        success            = "#2ca02c",
        error              = "#ff7f0e",
        skipped_preflight  = "#9467bd",
        skipped_filtering  = "#d62728",
        not_run            = "grey85"
    )

    combined$status <- factor(combined$status, levels = status_levels)
    combined$biosample <- factor(combined$biosample,
                                 levels = sort(unique(combined$biosample)))
    combined$contrast <- factor(combined$contrast,
                                levels = rev(sort(unique(combined$contrast))))
    combined$label <- ifelse(!is.na(combined$n_degs_base),
                             combined$n_degs_base, "")

    p <- ggplot(combined, aes(x = biosample, y = contrast, fill = status)) +
        geom_tile(color = "white", linewidth = 0.5) +
        geom_text(aes(label = label), size = 2.8, color = "black") +
        scale_fill_manual(values = status_colors, drop = FALSE) +
        labs(
            x = "Cell type", y = "Contrast", fill = "Status",
            title = "Pipeline Status Overview",
            subtitle = "Numbers = base DEGs (FDR < 0.05)"
        ) +
        theme_minimal(base_size = 11) +
        theme(
            axis.text.x = element_text(angle = 45, hjust = 1),
            panel.grid  = element_blank()
        )
    print(p)

    # Save to each contrast's summary dir
    for (cid in unique(combined$contrast)) {
        pdf_path <- file.path(cfg$outputs$results_dir, as.character(cid),
                              "summary", "status_tile.pdf")
        pdf(pdf_path, width = 14, height = 8)
        print(p)
        dev.off()
    }
    cat("Ō£ō Status tile plot saved\n")
}
```

## Final summary table

```{r final-summary}
if (exists("combined")) {
    cat("\nŌĢÉŌĢÉŌĢÉ Pipeline Summary ŌĢÉŌĢÉŌĢÉ\n\n")
    print(combined[, c("contrast", "biosample", "status",
                        "n_degs_base", "n_pathways_base")])
}
```

---

**Pipeline complete!** ­¤Äē

All results are in the `results/` directory. For deeper exploration,
use `notebooks/explore_results.ipynb`.
