---
title: "EasyDE Walkthrough — 07 Aggregate Results"
output: html_document
---

# Step 07: Aggregate Results

Runs once per contrast. Collects all per-stratum outputs and produces:
- `contrast_summary.csv` — one row per stratum with DEG counts and status
- `preflight_summary.csv` — preflight QC across all strata
- `status_tile.pdf` — visual overview of which strata succeeded/skipped
- Merged DE and fGSEA result tables across all strata
- Aggregated log summary

```{r setup, include=FALSE}
pipeline_root <- normalizePath("../..")
knitr::opts_knit$set(root.dir = pipeline_root)
setwd(pipeline_root)
config_path <- "config/config_treatments.yaml"   # EDIT THIS
```

## Source and run

```{r source-script}
source("workflow/scripts/07_aggregate_results.R")
```

```{r run-loop}
config_data <- load_config(config_path)
contrasts_df <- config_data$contrasts_df
cfg <- config_data$cfg

for (cid in unique(contrasts_df$contrast_id)) {
    cat(sprintf("\n=== Aggregating: %s ===\n", cid))
    tryCatch(
        main(config_path, cid),
        error = function(e) cat(sprintf("  [ERROR] %s\n", e$message))
    )
}
```

## View the summary

```{r view-summary}
for (cid in unique(contrasts_df$contrast_id)) {
    summary_path <- file.path(cfg$outputs$results_dir, cid,
                              "summary", "contrast_summary.csv")
    if (file.exists(summary_path)) {
        cat(sprintf("\n--- %s ---\n", cid))
        summary_df <- read.csv(summary_path, check.names = FALSE)
        print(summary_df[, c("biosample", "status", "n_degs_base",
                              "n_pathways_base")])
    }
}
```

## Status overview

```{r status-overview}
# Combine summaries across all contrasts
all_summaries <- list()
for (cid in unique(contrasts_df$contrast_id)) {
    summary_path <- file.path(cfg$outputs$results_dir, cid,
                              "summary", "contrast_summary.csv")
    if (file.exists(summary_path)) {
        all_summaries[[cid]] <- read.csv(summary_path, check.names = FALSE)
    }
}

if (length(all_summaries) > 0) {
    combined <- do.call(rbind, all_summaries)

    status_levels <- c("success", "error", "skipped_preflight",
                       "skipped_filtering", "not_run")
    status_colors <- c(
        success            = "#2ca02c",
        error              = "#ff7f0e",
        skipped_preflight  = "#9467bd",
        skipped_filtering  = "#d62728",
        not_run            = "grey85"
    )

    combined$status <- factor(combined$status, levels = status_levels)
    combined$biosample <- factor(combined$biosample,
                                 levels = sort(unique(combined$biosample)))
    combined$contrast <- factor(combined$contrast,
                                levels = rev(sort(unique(combined$contrast))))
    combined$label <- ifelse(!is.na(combined$n_degs_base),
                             combined$n_degs_base, "")

    options(repr.plot.width = 14, repr.plot.height = max(4, length(unique(combined$contrast)) * 1.2))

    p <- ggplot(combined, aes(x = biosample, y = contrast, fill = status)) +
        geom_tile(color = "white", linewidth = 0.5) +
        geom_text(aes(label = label), size = 2.8, color = "black") +
        scale_fill_manual(values = status_colors, drop = FALSE) +
        labs(
            x = "Cell type", y = "Contrast", fill = "Status",
            title = "Pipeline status (numbers = base DEGs)"
        ) +
        theme_minimal(base_size = 11) +
        theme(
            axis.text.x = element_text(angle = 45, hjust = 1),
            panel.grid  = element_blank()
        )
    print(p)
}
```

---

**Pipeline complete!** Check the `summary/` directory for all outputs.
Use the `explore_results.ipynb` notebook for deeper visualization.
